Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-06-05 doc/network.tex
one more line
tweak
stuff

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects. Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In Appendix, we present the same figure using smaller \RR (Appendix A1) and using Erlang-distributed latent periods (Appendix A2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

We compare two methods for estimating the effective generation-interval distribution and assume that the effective generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\RR$ \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption leads to slightly narrow confidence intervals, even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model) \swp{Appendix}.
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.

@@

To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
The SEIR model describes how a disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods. This gives periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}: 
\jd{Mention that our main-text models use exponential later instead.}
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{\sum_{k=1}^{n_I} I_k}{N}\\
\frac{dE_1}{dt} &= \beta S \frac{\sum_{k=1}^{n_I} I_k}{N} - n_E \sigma E_1\\
\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} I_n) && \text{for } n = 2, 3, \dots, n_I\\
\end{aligned}
\end{equation}
where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.

@@

\subsection{Comparison of estimates of reproductive number -- smaller $\RR$}
\pagebreak

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
This figure matches \fref{cmp} in the main text.
A smaller per-pair contact rate ($\lambda = 0.0026 \textrm{ days}^{-1}$) is used to simulate epidemics.
All other parameters are the same as in \fref{cmp}.
}
\label{fig:cmpsmall}
\end{figure}

@@

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
This figure matches \fref{cmp} in the main text.
Erland-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
All other parameters are the same as in \fref{cmp}.
}
\label{fig:cmpseminir}
\end{figure}

>>>>

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects. Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

We compare two methods for estimating the effective generation-interval distribution and assume that the effective generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\RR$ \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).

@@

To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
The SEIR model describes how a disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods. This gives periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}: 
\jd{Mention that our main-text models use exponential later instead.}
\swp{Done.}
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{\sum_{k=1}^{n_I} I_k}{N}\\
\frac{dE_1}{dt} &= \beta S \frac{\sum_{k=1}^{n_I} I_k}{N} - n_E \sigma E_1\\
\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} I_n) && \text{for } n = 2, 3, \dots, n_I\\
\end{aligned}
\end{equation}
where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
While we use the SEIR model for simulations in our main text, we show results based on more realistic distributions in Appendix.

@@

\subsection{Comparison of estimates of reproductive number -- smaller $\RR$}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive_small.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
This figure matches \fref{cmp} in the main text.
A smaller per-pair contact rate ($\lambda = 0.0026 \textrm{ days}^{-1}$) is used to simulate epidemics.
All other parameters are the same as in \fref{cmp}.
}
\label{fig:cmpsmall}
\end{figure}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive_seminr.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
This figure matches \fref{cmp} in the main text.
Erland-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
All other parameters are the same as in \fref{cmp}.
}
\label{fig:cmpseminir}
\end{figure}

\subsection{Testing the individual-based method on simulations on a homogeneous network}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../fig/full_coverage_fig.pdf}
\caption{\textbf{Wrong distributional assumption may result in biased estimates of the parameters of a generation-interval distribution.}
We simulate a stochastic SEIR model on a homogenous network with $10^5$ individuals using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$, mean infectious period $1/\gamma = 11.4 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$. Then, we apply the individual-based method based on the first 1000 infections.
(A-B) A boxplot of 100 estimates of the mean generation intervals and their coefficient of variations (CV). 
Dashed horizontal lines represent the true value.
(C) Coverage probability of the mean generation intervals and their CVs based on the 95\% confidence interval.
Dashed horizontal lines represent the 95\% coverage probability.
(D) Estimated generation-interval distributions based on the gamma approximation and a true generation-interval distribution of an SEIR model.
(E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$-$\RR$ relationship of an SEIR model.
Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}.
}
\label{fig:cover}
\end{figure}
